\artus -- Framework for Event-based Data Analysis in High Energy Physics

### 2015-09-23 Core/doc/latex/paper.tex
paper.tex: added top/bottom rule to table
paper.tex: one sentence per line for better version control
Complement labels and fix warnings

\usepackage{booktabs}
\usepackage{hepnames}
\usepackage{hepunits}
\usepackage{url}
\usepackage{palatino}
\usepackage[hidelinks=true,bookmarks=true,plainpages=false]{hyperref}
\usepackage[colorinlistoftodos,prependcaption]{todonotes}

@@

\section{Common Workflow of High Energy Physics Analyses \label{section_artus_analysis_workflow}}
% This section should contain a short description of a HEP/CMS-typical workflow.
% Especially common problems and challenges should be highlighted.
% From these problems, the design requirements for Artus are derived.
The workflow of the experiments collecting data at the Large Hadron Collider (LHC) at CERN is taken as an example to illustrate the organization and needs of a modern High Energy Physics (HEP) analysis. Still, equal or very similar requirements are also present for HEP experiments not located at the LHC.

@@

\section{Requirements and Challenges for Analysis Software\label{section_artus_challenges}}
%define requirements based on previous section, following sections: describe how they are fullfilled
Based on the workflow described in the previous section, the following demands for an analysis framework arise:

@@

\item[Maintainability:]
Source code maintenance, i.e. to be up-to-date with the latest requirements while not losing backwards compatibility, is a challenge for each software project. The first important step is to always have  an updated recipe that works out of the box on each machine offering CVMFS. To make the entry level as low as possible, beginners can start with an example analysis. In \artus, everything is steered via JSON configuration files. The modules then implement what has to be done while the data is stored in desired data format.

@@

\item[Reliability:]
In most cases, analyses never leave prototype status. This leads to frequent re-implementations and low trust in reliability. \artus allows the re-use of analysis code due to its modular structure as well as sharing it between different analysis groups, and therefore improving code quality significantly.

@@

\section{The \artus Framework \label{section_artus_motivation}}

@@

\section{Structure of the Framework \label{section_artus_structure}}

@@

\todo[inline]{I would not put the remaining part of this chapter here.
It should be either part of the requirements or part of the summary what Artus does achieve; but in this technical part in between it feels misplaced
-> Should we put this part into a dedicated section directly after this one?
I think the idea was to explain how certain design decisions, e.g. pipelines, achieve the \artus design goals.
Therefore it should not be too far from the technical descriptions.
}
This structure extends the established concepts of existing frameworks like FWLite~\cite{FWLite} from CMSSW to the following four main advantages:
\begin{description}
\item[Modular structure:] Analysis steps are implemented in processors, breaking down the complete analysis in well-defined and easily understandable and writable parts.
\item[Re-utilization of code:] Processors can be shared among different analysis groups.
Every user contributes to the testing of existing code and improves the confidence in this code.
\item[Configurability:] Each module uses settings that are read from the the configuration.
Each pipeline is mapped to a separate sub-set of the configuration.
\item[High performance:] The concept of pipelines allows to analyze the data with different configurations at the same time, while the data is only read in only once and common processing steps are shared in the global pipeline.
Therefore, the cost-intensive input operations are reduced to a minimum.
\end{description}

\section{Building an End-user Analysis \label{section_artus_analysis}}
\todo[inline]{If I would not know the structure already, the following part would not have helped my.
Either we support it with a drawing and more detailed explanations or we just refer to a readme...
Maybe rewrite parts of this section}
In the \artus framework, each concrete analysis is defined by a set of object types.
The analysis programmer must provide an implementation of the classes for settings, the event and the product class.
The settings class defines the tags that are read from the configuration file and the corresponding types of these values.
The event class defines the event content which is read from the input and the product class contains members for new quantities calculated during the analysis stage.
These three classes need to be derived from the base classes provided by the core of the \artus framework.

@@

\section{HarryPlotter -- Python Post-processing Framework \label{section_artus_harryplotter}}

@@

\section{A Real World Application -- the $\PH\to\Pgt\Pgt$ Analysis \label{section_artus_example_htt}}

@@

%time HiggsToTauTauAnalysis.py @HiggsAnalysis/KITHiggsToTauTau/data/ArtusWrapperConfigs/Sync13TeV.cfg -i HiggsAnalysis/KITHiggsToTauTau/data/Samples/NAF_sample_DYJetsToLLM50_RunIISpring15DR74_Asympt25ns_13TeV_MINIAOD_amcatnloFXFX-pythia8_recent.txt -f 10
\begin{table}[h!]
\centering
\begin{tabular}{lrr}
\textbf{Pipelines}                                & \textbf{Runtime / \minute} & \textbf{Event Rate / \reciprocal\second} \\ \midrule
%0                                                & 5:53                             & 1296 \\
 0                                                & 5:12                             & 1470 \\ \midrule
%\em                                              & 6:27                             & 1181 \\
 1 ($\Pe\Pe$)                                     & 6:31                             & 1174 \\
%1 ($\Pe\Pgt$)                                    & 7:03                             & 1084 \\
 1 ($\Pe\Pgt$)                                    & 7:26                             & 1028 \\
%1 ($\Pgm\Pgt$)                                   & 7:14                             & 1054 \\
 1 ($\Pgm\Pgt$)                                   & 7:14                             & 1052 \\
%1 ($\Pgt\Pgt$)                                   & 6:14                             & 1224 \\
 1 ($\Pgt\Pgt$)                                   & 6:07                             & 1247 \\ \midrule
%2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:33                             & 893 \\
 2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:32                             & 896 \\
%3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:59                             & 760 \\
 3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:54                             & 767 \\
%4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:46                            & 712 \\
 4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:01                            & 758 \\
\end{tabular}
\caption[Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.]{Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.
The analysis is performed on 10 files with a total size of 6.1~GB containing 452229 simulated $\PZ+\text{jet}$ events.
The measurement is performed with different numbers of pipelines, where each pipeline implements the analysis of one sub-channel of the analysis.
The values are compared with the execution without any pipelines and without global processors.
It is clearly visible, that the time needed for reading the inputs amounts to a large part of the overall runtime and that the simultaneous execution of multiple pipelines can reduce the runtime significantly compared to the case where different configuration are performed sequentially including separate IO operations.}
\label{table_artus_runtime_comparison}
\end{table}

@@

\section{Conclusion}
% what Artus does
% what it brought to the analysis:
% - code reuse among analysis groups
% - faster turnaround
% - faster new analysis
% - increased data in Run 2

>>>>

\usepackage{booktabs}
\usepackage{hepnames}
\usepackage{hepunits}
\usepackage{url}
\usepackage{palatino}
\usepackage[hidelinks,bookmarks=true,plainpages=false]{hyperref}
\usepackage[colorinlistoftodos,prependcaption]{todonotes}

@@

\section{Common Workflow of High Energy Physics Analyses}
\label{section_artus_analysis_workflow}
% This section should contain a short description of a HEP/CMS-typical workflow.
% Especially common problems and challenges should be highlighted.
% From these problems, the design requirements for Artus are derived.
The workflow of the experiments collecting data at the Large Hadron Collider (LHC) at CERN is taken as an example to illustrate the organization and needs of a modern High Energy Physics (HEP) analysis.
Still, equal or very similar requirements are also present for HEP experiments not located at the LHC.

@@

\section{Requirements and Challenges for Analysis Software}
\label{section_artus_challenges}
%define requirements based on previous section, following sections: describe how they are fullfilled
Based on the workflow described in the previous section, the following demands for an analysis framework arise:

@@

\item[Maintainability:]
Source code maintenance, i.e.\ to be up-to-date with the latest requirements while not losing backwards compatibility, is a challenge for each software project.
The first important step is to always have  an updated recipe that works out of the box on each machine offering CVMFS.
To make the entry level as low as possible, beginners can start with an example analysis.
In \artus, everything is steered via JSON configuration files.
The modules then implement what has to be done while the data is stored in desired data format.

@@

\item[Reliability:]
In most cases, analyses never leave prototype status.
This leads to frequent re-imple\-men\-ta\-tions and low trust in reliability.
\artus allows the re-use of analysis code due to its modular structure as well as sharing it between different analysis groups, and therefore improving code quality significantly.

@@

\section{The \artus Framework}
\label{section_artus_motivation}

@@

\section{Structure of the Framework}
\label{section_artus_structure}

@@

\todo[inline]{I would not put the remaining part of this chapter here.
It should be either part of the requirements or part of the summary what Artus does achieve; but in this technical part in between it feels misplaced
-> Should we put this part into a dedicated section directly after this one?
I think the idea was to explain how certain design decisions, e.g.\ pipelines, achieve the \artus design goals.
Therefore it should not be too far from the technical descriptions.
}
This structure extends the established concepts of existing frameworks like FWLite from CMSSW to the following four main advantages:
\begin{description}
\item[Modular structure:] Analysis steps are implemented in processors, breaking down the complete analysis in well-defined and easily understandable and writable parts.
\item[Re-utilization of code:] Processors can be shared among different analysis groups.
Every user contributes to the testing of existing code and improves the confidence in this code.
\item[Configurability:] Each module uses settings that are read from the the configuration.
Each pipeline is mapped to a separate sub-set of the configuration.
\item[High performance:] The concept of pipelines allows to analyze the data with different configurations at the same time, while the data is only read in only once and common processing steps are shared in the global pipeline.
Therefore, the cost-intensive input operations are reduced to a minimum.
\end{description}

\section{Building an End-user Analysis}
\label{section_artus_analysis}
\todo[inline]{If I would not know the structure already, the following part would not have helped my.
Either we support it with a drawing and more detailed explanations or we just refer to a readme...
Maybe rewrite parts of this section}
In the \artus framework, each concrete analysis is defined by a set of object types.
The analysis programmer must provide an implementation of the classes for settings, the event and the product class.
The settings class defines the tags that are read from the configuration file and the corresponding types of these values.
The event class defines the event content which is read from the input and the product class contains members for new quantities calculated during the analysis stage.
These three classes need to be derived from the base classes provided by the core of the \artus framework.

@@

\section{HarryPlotter -- Python Post-processing Framework}
\label{section_artus_harryplotter}

@@

\section{\texorpdfstring{A Real World Application -- the $\PH\to\Pgt\Pgt$ Analysis}{A Real World Application - the H to Tau Tau Analysis}}
\label{section_artus_example_htt}

@@

%time HiggsToTauTauAnalysis.py @HiggsAnalysis/KITHiggsToTauTau/data/ArtusWrapperConfigs/Sync13TeV.cfg -i HiggsAnalysis/KITHiggsToTauTau/data/Samples/NAF_sample_DYJetsToLLM50_RunIISpring15DR74_Asympt25ns_13TeV_MINIAOD_amcatnloFXFX-pythia8_recent.txt -f 10
\begin{table}[h!]
\centering
\begin{tabular}{lrr}
\toprule
\textbf{Pipelines}                                & \textbf{Runtime / \minute} & \textbf{Event Rate / \reciprocal\second} \\ \midrule
%0                                                & 5:53                             & 1296 \\
 0                                                & 5:12                             & 1470 \\ \midrule
%\em                                              & 6:27                             & 1181 \\
 1 ($\Pe\Pe$)                                     & 6:31                             & 1174 \\
%1 ($\Pe\Pgt$)                                    & 7:03                             & 1084 \\
 1 ($\Pe\Pgt$)                                    & 7:26                             & 1028 \\
%1 ($\Pgm\Pgt$)                                   & 7:14                             & 1054 \\
 1 ($\Pgm\Pgt$)                                   & 7:14                             & 1052 \\
%1 ($\Pgt\Pgt$)                                   & 6:14                             & 1224 \\
 1 ($\Pgt\Pgt$)                                   & 6:07                             & 1247 \\ \midrule
%2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:33                             & 893 \\
 2 ($\Pgm\Pgt$, $\Pe\Pgt$)                        & 8:32                             & 896 \\
%3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:59                             & 760 \\
 3 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$)             & 9:54                             & 767 \\
%4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:46                            & 712 \\
 4 ($\Pgm\Pgt$, $\Pe\Pgt$, $\Pe\Pgm$, $\Pgt\Pgt$) & 10:01                            & 758 \\
\bottomrule
\end{tabular}
\caption[Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.]{Measurement of the runtime of the baseline $\PH\to\Pgt\Pgt$ analysis.
The analysis is performed on 10 files with a total size of 6.1~GB containing 452229 simulated $\PZ+\text{jet}$ events.
The measurement is performed with different numbers of pipelines, where each pipeline implements the analysis of one sub-channel of the analysis.
The values are compared with the execution without any pipelines and without global processors.
It is clearly visible, that the time needed for reading the inputs amounts to a large part of the overall runtime and that the simultaneous execution of multiple pipelines can reduce the runtime significantly compared to the case where different configuration are performed sequentially including separate IO operations.}
\label{table_artus_runtime_comparison}
\end{table}

@@

\section{Conclusion}
\label{section_conclusion}
% what Artus does
% what it brought to the analysis:
% - code reuse among analysis groups
% - faster turnaround
% - faster new analysis
% - increased data in Run 2
